D-Hammer: Efficient Equational Reasoning for Labelled Dirac Notation

### 2025-04-06 02language_design.tex
.

\begin{lemma}[Soundness of equational theory]\label{lem: axiom sound}
If \( \Gamma \vdash e_1 = e_2 :T \) then \( \Gamma \vDash e_1 = e_2\).
\end{lemma}
% \begin{proof}
% By induction on the structure of the derivation
% \end{proof}
Next, we formalize the motivating example~\Cref{ex: motivating} in
plain Dirac notation.
\begin{example}[Motivating Example Formalization]
    \label{ex: formalizing motivating}
    Definitions and assumptions in the context \( \Gamma \) are formalized as follows:
    \begin{align*}
        & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
        &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
        & T && : \Index;  \qquad \qquad \qquad \qquad \qquad M \qquad : \OType(T, T).
    \end{align*}
    Notice how the functions and higher-order typing helps to formalize the abstract concepts here.
    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \). 
    The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the plain Dirac notation as:
    \[
    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
    \]
\end{example}

@@

Step 1 through 3 involve term rewriting for $R$. 
Term rewriting is the process of repeatedly reducing a term using a set of rules in the form of $l\ \reduce\ r$. The reduction works by matching the subterms with the left-hand side of a rule and replacing it with the right-hand side. 
For example, the term $(x\times y) . \ket{t} + \ket{t}$ is matched by the rule $a.K + K\ \reduce\ (a + 1).K$, and is rewritten into $(x\times y + 1) . \ket{t}$.
Step 1 and 3 use the same set of rewriting rules in~\Cref{sec: rewriting rules}.
Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$, which improves the decidability when dealing with sum.

Steps 4 through 6 are specialized algorithms designed to further normalize the axiom set $E$.
The main challenge here is the coexistence of AC-equivalence and SUM-SWAP, which means that naive sorting cannot alwasy convert equivalent terms into the same form.
For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted. Therefore we first ignore bound variables and normalize AC-equivalence by sorting. Afterward, the order of summation can be decided accordingly. The final step uses de Bruijn indices~\cite{deBruijn1972lambda} to resolve $\alpha$-equivalence. For further details, refer to~\Cref{sec: decide}.

>>>>

\begin{theorem}[Soundness of equational theory]\label{lem: axiom sound}
If \( \Gamma \vdash e_1 = e_2 :T \) then \( \Gamma \vDash e_1 = e_2\).
\end{theorem}
% \begin{proof}
% By induction on the structure of the derivation
% \end{proof}
Next, we formalize the motivating example~\Cref{ex: motivating} in
plain Dirac notation.
\begin{example}[Motivating Example Formalization]
    \label{ex: formalizing motivating}
    Definitions and assumptions in the context \( \Gamma \) are formalized as follows:
    \begin{align*}
        & \text{TPO} && := \lambda T_1 : \Index. \lambda T_2 : \Index. \lambda O : \OType(T_1, T_2). \sum_{i \in \mathbf{U}(T_1)} \sum_{j \in \mathbf{U}(T_2)} \bra{i} O \ket{j} . \ket{j}\bra{i} \\
        & &&\quad : \forall T_1. \forall T_2. \OType(T_1, T_2) \to \OType(T_2, T_1); \\
        &\text{phi} &&:= \lambda T : \Index. \sum_{i \in \mathbf{U}(T)} \sum_{j \in \mathbf{U}(T)} \ket{(i, j)} : \forall T.\KType(T \times T); \\
        & T && : \Index;  \qquad \qquad \qquad \qquad \qquad M \qquad : \OType(T, T).
    \end{align*}
    Notice how the functions and higher-order typing helps to formalize the abstract concepts here.
    The symbol \( \text{TPO} \) represents the transpose of an operator, polymorphic on the Hilbert spaces \( T_1 \) and \( T_2 \). 
    The symbol \( \text{phi} \) takes the index \( T \) and defines the maximally entangled states, summing over all basis elements in \( T \), as indicated by the universal set \( \mathbf{U}(T) \).
    With the assumption of the index \( T \) and operator \( M \), we can express the equivalence in the plain Dirac notation as:
    \[
    (\textrm{M} \otimes \mathbf{1}_\mathcal{O}(\textrm{T})) \cdot (\textrm{phi T}) = (\mathbf{1}_\mathcal{O}(\textrm{T}) \otimes (\textrm{TPO T T M})) \cdot (\textrm{phi T}).
    \]
\end{example}

@@

Step 1 through 3 involve term rewriting for $R$. 
Term rewriting is the process of repeatedly reducing a term using a set of rules in the form of $l\ \reduce\ r$. The reduction works by matching the subterms with the left-hand side of a rule and replacing it with the right-hand side. 
For example, the term $(x\times y) . \ket{t} + \ket{t}$ is matched by the rule $a.K + K\ \reduce\ (a + 1).K$, and is rewritten into $(x\times y + 1) . \ket{t}$.
Step 1 and 3 use the same set of rewriting rules in~\Cref{sec: rewriting rules}.
Step 2 expands variables to their abstract element-wise representation, e.g., $K \ \reduce\ \sum_{i} (\bra{i} \cdot K). \ket{i}$, which is useful when reasoning about sums.

Steps 4 through 6 are specialized algorithms designed to further normalize the axiom set $E$.
The main challenge here is the coexistence of AC-equivalence and SUM-SWAP, which means that naive sorting cannot alwasy convert equivalent terms into the same form.
For step 4 and 5, the key observation is that in a successive sum expression \( \sum_{i \in s_1} \cdots \sum_{j \in s_n} A \), the names and order of the bound variables \( i, \dots, j \) can be freely permuted. Therefore we first ignore bound variables and normalize AC-equivalence by sorting. Afterward, the order of summation can be established accordingly. The final step uses de Bruijn indices~\cite{deBruijn1972lambda} to resolve $\alpha$-equivalence. For further details, refer to~\Cref{sec: decide}.

### 2025-04-06 06related_conclusion.tex
.

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,vandewetering2020zx}
is a popular graphical calculus for reasoning about quantum states.
The foundations of the ZX calculus are grounded in categorical quantum
mechanics (CQM)~\cite{DBLP:conf/lics/AbramskyC04}, which provides an
category-theoretic foundation for quantum mechanics.  A key aspect of
the ZX calculus and related graphical calculi is that they support a
form of graph rewriting. Two notable properties of these rewriting
systems are:
\begin{itemize}
\item completeness:
\item efficiency:
\end{itemize}
Both Labelled Dirac Notation (LDN) and the ZX calculus share the use
of rewriting techniques. However, it is difficult to compare the ZX
calculus and LDN. First, they have different scopes: the ZX calculus
mainly focuses on quantum circuits, whereas LDN aims to provide
general support for quantum physics. This difference in scope is
reflected in the underlying languages: for instance, many works in ZX
calculus consider concrete sets of quantum gates, whereas LDN
emphasizes the use of big operators, and does not attempt to provide
specialized support for a particular set of quantum gates. As a
result, many of the examples handled by D-Hammer cannot be formalized
in existing ZX-calculus tools. On the contrary, D-Hammer is not
competitive on the examples that can be verified by ZX-calculus.

>>>>

\paragraph*{Comparison with ZX Calculus}
The ZX calculus~\cite{DBLP:conf/icalp/CoeckeD08,coecke2011interacting}
is a graphical calculus for quantum states. A main appeal of the ZX
calculus is that its foundations are grounded in categorical quantum
mechanics~\cite{DBLP:conf/lics/AbramskyC04}, a powerful framework for
modeling and reasoning about quantum physics. Another main appeal of
the ZX calculus is that ZX diagrams have an operational interpretation
using graph rewriting. There is a large body of work that studies
properties of these rewriting systems, and in particular completeness
(a set of rules is complete if it can prove all valid identities) and
minimality (a complete set of rules is minimal if removing a rule
leads to incompleteness). Two main proof techniques for completeness
are via termination, or via interpretation. In the first case, one
shows that a rewriting system has unique normal forms and that two
expressions are semantically equivalent iff they have the same normal
form---there exists many relaxations of this result---whereas in the
second case one shows completeness by exhibiting a well-behaved
translation to another system for which completeness holds. Both proof
techniques have been used to prove completeness for multiple settings,
including the Clifford fragment~\cite{backens2014zx}, the
Toffoli+Hadamard fragment~\cite{hadzi15zx}, the Clifford+T
fragment~\cite{jeandel2018complete} and the (qubit) universal
fragment~\cite{HNW18,JeandelPV18beyond}. Subsequent works generalize
completeness to qudits~\cite{Poor2023} for a calculus know as ZXW
which encompasses the ZX and ZW calculi~\cite{zwcalculus}. We refer
the reader to~\cite{vandewetering2020zx} for a (relatively) recent
historical and technical account of completeness results.

Because they are both grounded in rewriting techniques, there are many
similarities between prior work on the ZX calculus and our own work on
labelled Dirac notation. We highlight some key differences. First,
there exist a number of differences between our language and the
concrete fragments of the ZX-calculus for which completeness has been
studied. For instance, our language features big operators, and
operators for dual spaces. A second difference is that we do not claim
completeness: our main theorem shows a reduction from Labelled Dirac
notation to Dirac notation, for which completeness is only partially
known.

it is difficult to compare the ZX
calculus and LDN. First, they have different scopes: the ZX calculus
mainly focuses on quantum circuits, whereas LDN aims to provide
general support for quantum physics. This difference in scope is
reflected in the underlying languages: for instance, many works in ZX
calculus consider concrete sets of quantum gates, whereas LDN
emphasizes the use of big operators, and does not attempt to provide
specialized support for a particular set of quantum gates. As a
result, many of the examples handled by D-Hammer cannot be formalized
in existing ZX-calculus tools. On the contrary, D-Hammer is not
competitive on the examples that can be verified by ZX-calculus.
